### 2015-07-22 Harris_2015/manuscript-materials/draft.Rmd
draft

In this paper, following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I show that Markov networks [undirected graphical models
also known as Markov random fields; @murphy_machine_2012] can provide a framework
for understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observational data. Markov networks
have been used in many scientific fields for decades to model associations
between various kinds of "particles". For example, a well-studied network named the Ising model has played an
important role in our understanding of physics [where nearby particles tend to
align magnetically with one another; @cipra_introduction_1987]. In spatial
contexts, these models have been used to describe interactions between
adjacent grid cells [@harris_contact_1974; @gelfand_modelling_2005]. In
neurobiology, they have helped researchers determine which neurons are
connected to one another by modeling the structure in their firing patterns
[@schneidman_weak_2006].

Doing so
would allow ecologists to simulate and study the landscape-level consequences
of arbitrary species interaction matrices, even when our observations are not
Gaussian. While ecologists explored some related approaches in the 1980's
[@whittam_species_1981], computational limitations had previously imposed
severe approximations that produced unintelligible results [e.g.
"probabilities" greater than one; @gilpin_factors_1982].  Now that it is
computationally feasible to fit these models exactly, the approach has become
worth a second look.

@@

Estimating the marginal relationships predicted by a Markov network is more
difficult than estimating conditional relationships, because doing so requires
absolute probability estimates. Turning the relative probability given by
Equation 1 into an absolute probability entails scaling by a *partition
function*, $Z(\alpha, \beta)$, which is defined so that that the probabilities of all
possible assemblages that could be produced by the model sum to one (bottom of
Figure 2B). Calculating $Z(\alpha, \beta)$ exactly, as is done in this paper,
quickly becomes infeasible as the number of species increases: with
$2^N$ possible assemblages of $N$ species, the number of bookkeeping operations
required for exact estimation spirals exponentially into the billions and beyond.
Numerous techniques are available for working with Markov networks that keep
the computations tractable, e.g. via analytic approximations
[@lee_learning_2012] or Monte Carlo sampling [@salakhutdinov_learning_2008],
but they are beyond the scope of this paper.

@@

For analyses with 5 or more species, I made a small modification to the maximum
likelihood procedure described above. Given the large number of parameters
associated with some of the networks to be estimated, I regularized the
likelihood using a logistic prior distribution [@gelman_weakly_2008] with
a scale of 1 on the $\alpha$ and $\beta$ terms.

\noindent  
***Comparison with existing techniques.***
After fitting Markov networks to the simulated landscapes described above, I
used several other techniques for determining the sign and strength of marginal
associations between pairs of species (Appendix B).

The first two alternative interaction measures were the sample covariances and the
partial covariances between each pair of species' data vectors on the landscape
[@albrecht_spatial_2001]. Because partial covariances are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on ridge regression [@van_wieringen_ridge_2014; i.e. linear
regression with a Gaussian prior]. For these analyses, I set the ridge
parameter to 0.2 divided by the number of sites on the landscape.

The third alternative method, described in @gotelli_empirical_2009, involved simulating
possible landscapes from a null model that retains the row and column sums of
the original matrix [@strong_ecological_1984]. Using the default options in the
Pairs software described in @gotelli_empirical_2009, I simulated  the null
distribution of scaled C-scores (a test statistic describing the number of
*non*-co-occurrences between two species). The software then calculated a
$Z$ statistic for each species pair using this null distribution. After
multiplying this statistic by $-1$ so that positive values corresponded to
facilitation and negative values corresponded to competition, I used it as
another estimate of species interactions.

\noindent  
***Method evaluation.***
For the simulated landscapes based on Figure 1, method evaluation was fairly
qualitative: any method whose test statistic for the two shrubs indicated a
negative relationship passed; other methods failed.

For the larger landscapes, I rescaled the four methods' estimates using linear
regression through the origin so that they all had a consistent
interpretation. In each method, I regressed the "true" $\beta$ coefficient for
each species pair against the model's estimate, re-weighting the pairs so that
each landscape contributed equally to the rescaled estimate. These regressions
yielded squared errors for each of the 23,140 simulated species pairs, across
all conditions and replicates. I then partitioned these errors into groups
defined by the properties of the simulated landscapes (i.e. species richness,
number of observed communities, and the presence/absence of environmental
filtering). Within each partition, the mean squared error was used to
calculate the proportion of variance explained (compared with a baseline model
that assumed all interaction strengths to be zero).

>>>>

In this paper, following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I show that Markov networks [undirected graphical models
also known as Markov random fields; @murphy_machine_2012] can provide a framework
for understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observational data. Markov networks
have been used in many scientific fields for decades to model associations
between various kinds of "particles". For example, a well-studied network named the Ising model has played an
important role in our understanding of physics [where nearby particles tend to
align magnetically with one another; @cipra_introduction_1987]. In spatial
contexts, these models have been used to describe interactions between
adjacent grid cells [@harris_contact_1974; @gelfand_modelling_2005]. In
neurobiology, they have helped researchers determine which neurons are
connected to one another by modeling the structure in their firing patterns
[@schneidman_weak_2006]. While ecologists explored some related approaches in the 1980's
[@whittam_species_1981], computational limitations had previously imposed
severe approximations that produced unintelligible results [e.g.
"probabilities" greater than one; @gilpin_factors_1982].  Now that it is
computationally feasible to fit these models exactly, this kind of approach has become
worth a second look.

@@

Estimating the marginal relationships predicted by a Markov network is more
difficult than estimating conditional relationships, because doing so requires
absolute probability estimates. Turning the relative probability given by
Equation 1 into an absolute probability entails scaling by a normalizing constant, $Z(\alpha, \beta)$, which is defined so that that the probabilities of all
possible assemblages that could be produced by the model sum to one (bottom of
Figure 2B). Calculating $Z(\alpha, \beta)$ exactly, as is done in this paper,
quickly becomes infeasible as the number of species increases: with
$2^N$ possible assemblages of $N$ species, the number of bookkeeping operations
required spirals exponentially into the billions and beyond.
Numerous techniques are available for working with Markov networks that keep
the computations tractable, e.g. via analytic approximations
[@lee_learning_2012] or Monte Carlo sampling [@salakhutdinov_learning_2008],
but they are beyond the scope of this paper.

@@

\noindent  
***Recovering species interactions from simulated data.***
I used Markov networks and  several other techniques for determining the sign and strength
of the associations between pairs of species (Appendix B).

For the Markov networks, I used the rosalia package [@harris_rosalia_2015], as described
above. Given the large number of parameters associated with some of the networks to
be estimated, I regularized the likelihood using a Cauchy prior distribution
[@gelman_weakly_2008] with a scale of 1 on the $\alpha$ and $\beta$ terms.

I also evaluated five alternative methods from the existing literature to calibrate the
performance of the Markov network as a method for estimating species interactions.
The first two alternative interaction measures were the sample correlations and the
partial correlations between each pair of species' data vectors on the landscape
[@albrecht_spatial_2001; @faisal_inferring_2010]. Because partial correlations are undefined
for landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
`pcor.shrink` function with the default settings [@REF]. In the context of non-Gaussian
data, the partial correlation can be thought of as an approximation to the Markov network,
and is much easier to compute.

The third alternative, generalized linear models, can also be thought of as a computationally
efficient approximation to the Markov network [@REF].  Following @faisal_inferring_2010,
I fit regularized logistic regression models for each species, using the other species on the
landscape as predictors. [[Observational presence-absence data does not have enough degrees
of freedom to estimate this many parameters, so I symmetrized the pairwise correlations by
averaging the coefficient for predicting species $i$’s status from species $j$ and vice versa.]]

The fourth alternative method, described in @gotelli_empirical_2009, involved simulating
possible landscapes from a null model that retains the row and column sums of
the original matrix [@strong_ecological_1984]. Using the default options in the
Pairs software described in @gotelli_empirical_2009, I simulated  the null
distribution of scaled C-scores (a test statistic describing the number of
*non*-co-occurrences between two species). The software then calculated a
$Z$ statistic for each species pair using this null distribution. After
multiplying this statistic by $-1$ so that positive values corresponded to
facilitation and negative values corresponded to competition, I used it as
another estimate of species interactions.

Finally, I used the latent correlation matrix estimated by the BayesComm package [@REF]
to evaluate the claim from recent papers that the correlation coefficients estimated by “joint
species distribution models” provide an accurate assessment of species’ pairwise interactions
[@pollock_understanding_2014; see also @harris_generating_2015].

\noindent  
***Evaluating model performance.***
For the simulated landscapes based on Figure 1, method evaluation was fairly
qualitative: any method whose test statistic for the two shrubs indicated a
negative relationship passed; other methods failed. For methods that provide
confidence intervals or other inferential statistics, I reported them below.

Because the different methods mostly describe species interactions on different
scales (e.g. correlations versus $Z$ scores versus regression coefficients), I used
linear regression through the origin to rescale the different estimates produced
by each method so that they had a consistent interpretation. After rescaling each
method’s estimates, I calculated squared errors for each of the [[N]] species pairs
across all the simulated data sets and used these values to calculate the proportion of variance
explained for landscapes with 25, 200, or 1600 sampled locations (compared with a baseline
model that assumed all interaction strengths to be zero).
